---
title: "Calculating anthropometric z-scores using zscorer"
author: "Mark Myatt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating anthropometric z-scores using zscorer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if(!require(zscorer)) install.packages("zscorer")

```

# Calculating anthropometric z-scores using `zscorer`

The main function in the `zscorer` package is `addWGSR()`.

To demonstrate its usage, we will use the accompanying dataset in `zscorer` called `anthro3`. We inspect the dataset as follows:

```{r usage1, echo = TRUE, eval = FALSE}
head(anthro3)
```

which returns:

```{r usage1a, echo = FALSE, eval = TRUE}
head(anthro3)
```

`anthro3` contains anthropometric data from a Rapid Assessment Method (RAM) survey from Burundi.

Anthropometric indices (e.g. weight-for-height z-scores) have not been calculated and added to the data.

We will use the `addWGSR()` function to add weight-for-height (wfh) z-scores to the example data:

```{r usage2, echo = TRUE, eval = TRUE}
svy <- addWGSR(data = anthro3, sex = "sex", firstPart = "weight",
               secondPart = "height", index = "wfh")
```

A new column named **wfhz** has been added to the dataset:

```{r usage2a, echo = FALSE, eval = TRUE}
head(svy)
```

The `wfhz` column contains the weight-for-height (wfh) z-scores calculated from the `sex`, `weight`, and `height` columns in the `anthro3` dataset. The calculated z-scores are rounded to two decimals places unless the `digits` option is used to specify a different precision (run `?addWGSR` to see description of various parameters that can be specified in the `addWGSR()` function). 

The `addWGSR()` function takes up to nine parameters to calculate each index separately, depending on the index required. These are described in the *Help* files of the `zscorer` package which can be accessed as follows:

```{r usage2b, echo = TRUE, eval = FALSE}
?addWGSR
```

The **standing** parameter specifies how “stature” (i.e. length or height) was measured. If this is not specified, and in some special circumstances, height and age rules will be applied when calculating z-scores. These rules are described in the table below.

&nbsp;

+---------------+---------------+---------------+---------------+----------------------------------------+
| **index**     | **standing**  | **age**       | **height**    | **Action**                             |
+===============+===============+===============+===============+========================================+
| hfa or lfa    | standing      | < 731 days    |               | index = lfa                            |
|               |               |               |               | height = height + 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+
| hfa or lfa    | supine        | < 731 days    |               | index = lfa                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| hfa or lfa    | unknown       | < 731 days    |               | index = lfa                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| hfa or lfa    | standing      | ≥ 731 days    |               | index = hfa                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| hfa or lfa    | supine        | ≥ 731 days    |               | index = hfa                            |
|               |               |               |               | height = height - 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+
| hfa or lfa    | unknown       | ≥ 731 days    |               | index = hfa                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | standing      |               | < 65 cm       | index = wfl                            |
|               |               |               |               | height = height + 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | standing      |               | ≥ 65 cm       | index = wfh                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | supine        |               | ≤ 110 cm      | index = wfl                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | supine        |               | more than     | index = wfh                            |
|               |               |               | 110 cm        | height = height - 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | unknown       |               | < 87 cm       | index = wfl                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| wfh or wfl    | unknown       |               | ≥ 87 cm       | index = wfh                            |
+---------------+---------------+---------------+---------------+----------------------------------------+
| bfa           | standing      | < 731 days    |               | height = height + 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+
| bfa           | standing      | ≥ 731 days    |               | height = height - 0.7 cm               |
+---------------+---------------+---------------+---------------+----------------------------------------+

&nbsp;


The `addWGSR()` function will not produce error messages unless there is something very wrong with the data or the specified parameters. If an error is encountered in a record then the value **NA** is returned. Error conditions are listed in the table below. 

&nbsp;

+--------------------------------------------------+----------------------------------------+
| **Error condition**                              | **Action**                             |
+==================================================+========================================+
| Missing or nonsense value in `standing` parameter| Set `standing` to `3` (unknown) and    | 
|                                                  | apply appropriate height or age rules. |
+--------------------------------------------------+----------------------------------------+
| Unknown `index` specified                        | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| Missing `sex`                                    | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| Missing `firstPart`                              | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| Missing `secondPart`                             | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| `sex` is not male (`1`) or female (`2`)          | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| `firstPart` is not numeric                       | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| `secondPart` is not numeric                      | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| Missing `thirdPart` when `index = "bfa"`         | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| `thirdPart` is not numeric when `index = "bfa"`  | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
| `secondPart` is out of range for specified index | Return **NA** for z-score.             |
+--------------------------------------------------+----------------------------------------+
 
&nbsp;

We can see this error behaviour using the example data:

```{r usage3, echo = TRUE, eval = TRUE}
table(is.na(svy$wfhz))
```

We can display the problem record:

```{r usage4, echo = TRUE, eval = TRUE}
svy[is.na(svy$wfhz), ]
```

The problem is due to the value **9** in the `sex` column, which should be coded **1** (for male) and **2** (for female). Z-scores are only calculated for records with sex specified as either **1** (male) or **2** (female). All other values, including **NA**, will return **NA**.

The `addWGSR()` function requires that data are recorded using the required units or required codes (see Table Z1).

The `addWGSR()` function will return incorrect values if the data are not recorded using the required units. For example, this attempt to add weight-for-age z-scores to the example data:

```{r usage5, echo = TRUE, eval = TRUE}
svy <- addWGSR(data = svy, sex = "sex", firstPart = "weight", 
               secondPart = "age", index = "wfa")
```
	
will give incorrect results:

```{r usage5a, echo = TRUE, eval = TRUE}
summary(svy$wfaz)
```

The odd range of values is due to age being recorded in months rather than days.

It is simple to convert all ages from months to days:

```{r usage5b, echo = TRUE, eval = TRUE}
svy$age <- svy$age * (365.25 / 12)
head(svy)
```

before calculating and adding weight-for-age z-scores:

```{r usage5c, echo = TRUE, eval = TRUE}
svy <- addWGSR(data = svy, sex = "sex", firstPart = "weight", 
               secondPart = "age", index = "wfa")
head(svy)
summary(svy$wfaz)
```

The muac column in the example dataset is recorded in millimetres (mm). We need to convert this to centimetres (cm): 

```{r usage6, echo = TRUE, eval = TRUE}
svy$muac <- svy$muac / 10
head(svy)
```

before using the `addWGS()` function to calculate MUAC-for-age z-scores:

```{r usage6a, echo = TRUE, eval = TRUE}
svy <- addWGSR(svy, sex = "sex", firstPart = "muac",	
               secondPart = "age", index = "mfa")
head(svy)
```

As a last example we will use the `addWGSR()` function to add body mass index-for-age (bfa) z-scores to the data to create a new variable called bmiAgeZ with a precision of 4 decimal places as:

```{r usage7, echo = TRUE, eval = TRUE}
svy <- addWGSR(data = svy, sex = "sex", firstPart = "weight", 
               secondPart = "height", thirdPart = "age", index = "bfa", 
               output = "bmiAgeZ", digits = 4)
head(svy)
```

## Calculating z-scores using the legacy functions

The `zscorer` package comes with the original legacy functions included in its `version 0.1.0`. These functions allow for the calculation of `weight-for-age`,  `height-for-age` and `weight-for-height z-scores` for individual children and for a cohort of children.

### Calculating z-score for each of the three anthropometric indices for a single child

For this example, we will use the `getWGS()` function and apply it to dummy data of a **52 month** old male child with a weight of **14.6 kg** and a height of **98.0 cm**.

```{r example1, eval = TRUE}
# weight-for-age z-score
waz <- getWGS(sexObserved = 1,     # 1 = Male / 2 = Female
              firstPart = 14.6,    # Weight in kilograms up to 1 decimal place
              secondPart = 52,     # Age in whole months
              index = "wfa")       # Anthropometric index (weight-for-age)

waz

# height-for-age z-score
haz <- getWGS(sexObserved = 1,
              firstPart = 98,      # Height in centimetres
              secondPart = 52,
              index = "hfa")       # Anthropometric index (height-for-age)

haz

# weight-for-height z-score
whz <- getWGS(sexObserved = 1,
              firstPart = 14.6,
              secondPart = 98,
              index = "wfh")       # Anthropometric index (weight-for-height)

whz
```

Applying the `getWGS()` function results in a calculated `z-score` for one child.

### Calculating z-score for each of the three anthropometric indices for a cohort or sample of children

For this example, we will use the `getCohortWGS()` function and apply it to sample data `anthro1` that came with `zscorer`.

```{r sample-data1, eval = FALSE}
# Make a call for the anthro1 dataset
anthro1
```

As you will see, this dataset has the 4 variables you will need to use with `getCohortWGS()` to calculate the `z-score` for the corresponding anthropometric index. These are `age`, `sex`, `weight` and `height`.

```{r, echo = FALSE, eval = TRUE}
library(zscorer)
```

```{r sample-data2, eval = TRUE}
head(anthro1)
```

To calculate the three anthropometric indices for all the children in the sample, we execute the following commands in R:

```{r example2, eval = TRUE}
# weight-for-age z-score
waz <- getCohortWGS(data = anthro1,
                    sexObserved = "sex",
                    firstPart = "weight",
                    secondPart = "age",
                    index = "wfa")
head(waz, 50)

# height-for-age z-score
haz <- getCohortWGS(data = anthro1,
                    sexObserved = "sex",
                    firstPart = "height",
                    secondPart = "age",
                    index = "hfa")
head(haz, 50)

# weight-for-height z-score
whz <- getCohortWGS(data = anthro1,
                    sexObserved = "sex",
                    firstPart = "weight",
                    secondPart = "height",
                    index = "wfh")
head(whz, 50)
```

Applying the `getCohortWGS()` function results in a vector of calculated `z-scores` for all children in the cohort or sample.

### Calculating z-scores for all of the three anthropometric indices in one function

For this example, we will use the `getAllWGS()` function and apply it to sample data `anthro1` that came with `zscorer`.

```{r example3, eval = TRUE}
# weight-for-age z-score
zScores <- getAllWGS(data = anthro1,
                     sex = "sex",
                     weight = "weight",
                     height = "height",
                     age = "age",
                     index = "all")
head(zScores, 20)
```

Applying the `getAllWGS()` function results in a data frame of calculated `z-scores` for all children in the cohort or sample for all the anthropometric indices.
